Working title: Verifying the Edmonds-Karp Algorithm

### 2016-02-29 paper/root.tex
not compiling
Darfts of impl/refine sections

\pagestyle{plain} % turn on page numbers
\usepackage[utf8]{inputenc}
\usepackage{microtype} % Better typesetting for PDFs -- is enabling this ok?
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{eufrak} %The eufrak package is redundant if the amsfonts package is used
% \usepackage{mathpartir}
\DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}
\usepackage[boxed]{algorithm}
\usepackage{enumerate}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{color}
\usepackage[noend]{algpseudocode}
\usepackage{caption}
\usepackage[font=scriptsize]{subcaption}
\usepackage{hyperref}
\usepackage{float}
\usepackage{wrapfig}

@@

@@

@@

@@

  Introduction to refinement. (On abstract level)

  TODO: Show better complexity bound for FoFu with shortest path! --> Edmonds-Karp algo!

@@

  Down to executable code:
    Implement things that where not yet specified/specified in a non-executable way:
      Residual graph: By combination of c and f, and pred-succ. TODO: Directly!
        All we need is successor function. Implemented by tabulating 
          adjacent nodes of each node (pred-succ), and using c and f to filter out actual successor nodes. 
      augpath-spec by BFS/DFS (which work on succ-functions)
      bottleneck and augment: Give iterative impl.

    Use efficient data structures:
      Arrays for c,f, and ps. c and f in row-major indexing.

@@

\section{Benchmarking}
\section{Conclusion} 
  ... and related work
    Mizar-Formalization: What did they do.

@@

  Formal proof of mincut maxflow
  fofo-scheme
  inst to edmonds karp
  refinement down to executable code. Roughly 5 times slower than Java. (What about OCaml)
    + NetCheck

>>>>

\pagestyle{plain} % turn on page numbers
\usepackage[utf8]{inputenc}
\usepackage{microtype} % Better typesetting for PDFs -- is enabling this ok?
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{eufrak} %The eufrak package is redundant if the amsfonts package is used
% \usepackage{mathpartir}
%\DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}
\usepackage[boxed]{algorithm}
\usepackage{enumerate}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{color}
\usepackage[noend]{algpseudocode}
\usepackage{caption}
\usepackage[font=scriptsize]{subcaption}
\usepackage{hyperref}
\usepackage{float}
\usepackage{wrapfig}

@@

\usepackage{packages/isabelle}
\usepackage{packages/isabelletags}
\usepackage{packages/isabellesym}
\usepackage{packages/comment}

@@

% Include snippets
\newcommand{\DefineSnippet}[2]{%
   \expandafter\newcommand\csname snippet--#1\endcsname{%
     \begin{quote}
     \begin{isabelle}
     #2
     \end{isabelle}
     \end{quote}}}
\newcommand{\Snippet}[1]{\csname snippet--#1\endcsname}
\input{snippets/snippets}

@@

\Snippet{isar-lem}

\section{Refinement in Isabelle}
Purpose: Separation of concerns.

Short intro on Refinement Framework:

Programs phrased as error-set monad. With order structure. 
Refinement: Less possible results (refinement ordering). Error is greatest element (refined by everything).

Building chains of refinement, at the end there should be a deterministic algorithm (yielding exactly one result), which is
contained (by transitivity) in the specification.

Standard combinators (bind, recursion, if, etc) are monotonic wrt. refinement ordering, which allows for modularity:
Replacing part of a program by a refined version, yields a refined program.

Data refinement: Changing the representation of data. Relation between concrete and abstract representation. 
$\Uparrow m$ is biggest concrete program that refines $m$ wrt.\ relation $R$.

The Isabelle Refinement Framework provides the nres-monad, and tools to reason about program refinement, 
the most important one being a VCG.

\section{Edmonds-Karp Algorithm}
  The general Ford-Fulkerson method can only be shown to terminate if the edge capacities are integer numbers. 
  An improvement over that is the Edmonds-Karp algorithm~\cite{}, which is obtained when always choosing 
  a shortest augmenting path. In this case, it can be shown that only $O(|V||E|)$ augmentations are performed until the
  algorithm terminates, even for real-valued capacities. A shortest augmenting path can be obtained by breadth first search in time $O(|E|)$,
  yielding an overall complexity of $O(|V||E|^2) $ algorithm.

  In our formalization, we refine the specification of an augmenting path to a specification of a shortest augmenting path, which immediately yields 
  an abstract version of the Edmonds-Karp algorithm which refines the Ford-Fulkerson method.

  The larger part of the formalization is spent on proving the complexity bound. Note that the refinement framework does not 
  have a notion of computational complexity, so we cannot even define the runtime of an algorithm. However, we can instrument 
  the while-loop of the algorithm with a counter, which is incremented on each iteration, and prove an upper bound on this counter.
  Moreover, this also yields a termination argument for real-valued capacities.\footnote{As we restricted the algorithm 
  to integer valued capacities at the very abstract level, this is actually not used.}

  The idea of the complexity proof is as follows: Note that edges that are reverse to edges on a shortest augmenting path cannot lie 
  on a shortest path itself.
  On every augmentation, at least one edge that lies on a shortest path is flipped. Thus, either the length of the shortest path increases,
  or the number of edges that lie on a shortest path decreases. As the length of a shortest path is at most $|V|$, there are no more than $O(|E||V|)$ iterations.

  Formalizing the above intuitive argument is more tricky than it seems on first glance. 
  While it easy to prove that, in a \emph{fixed graph}, an edge and its reverse cannot both lie on shortest paths, generalizing the argument
  to a graph transformation which may add reverse edges and removes at least original edge, is tricky. Note that a straightforward induction
  on the length of the augmenting path must fail, as after flipping the first edge, the path is no longer augmenting.

  We decided to generalize the statement as follows. Assume we have an original graph $cf$ with a shortest augmenting path $p$, and a transformed graph $cf'$ that has been created from
  $cf$ by adding some flipped edges from $p$ and removing some edges from $p$.
  Then, we consider an edge $(u,v) \in p$, a path $p_1$ from $s$ to $v$ in the original graph, and a path $p_2$ from $u$ to $t$ in the transformed graph.

  By induction on the number of reversed edges in $p_2$, we 
  show that $|p_1|+|p_2| > |p|$. In the induction step, we use the proof idea for the single flipped edge $(v,u)$, and,
  if $p_2$ contains more flipped edges, we split $p_2$ at the first flipped edge. Then, the initial segment of $p_2$ contains no flipped edges,
  and thus also is a path in the original graph, and we can apply the induction hypothesis.

  Proving that augmentation with a shortest augmenting path actually only adds flipped edges, and removes at least one original edge, required some more work,
  but was straightforward and yielded no surprises.

  Finally, we phrase the complexity argument as a measure function $m = l*2|E| + n$, where $l$ is the length of the shortest augmenting path,
  and $n$ is the number of edges that lie on any shortest path. We show that $m$ is decreased by the loop body.
  Adding a special case for loop termination (there are no augmenting paths), and observing that $m$ is bounded by $2|V||E|$, yields 
  the desired upper bound on loop iterations. Refinement of the algorithm to add an explicit loop counter, and asserting the upper bound, is then straightforward.

@@

  In the previous sections, we have described the Ford-Fulkerson and the Edmonds-Karp algorithm abstractly,  
  leaving open how to obtain a (shortest) augmenting path. A standard way to find a shortest path in a graph 
  is breadth first search (BFS). Luckily, we had already formalized a BFS algorithm as an example for the refinement framework.
  However, the existing formalization could only compute the minimum distance between two nodes, without returning an actual path.
  We briefly report on our adapted formalization here, which is displayed in Figure~\fig{TODO: Snippet}: 
  TODO: Reference to figure (line numbers?)
  The abstract algorithm keeps track of a set $C$ of current nodes, a set $N$ of new nodes, and a partial map $P$ from already visited nodes to predecessor nodes.
  Initially, only the start node is in $C$ and $N$ is empty. In each iteration of the loop, a node $u\in C$ is picked, and its successors $v$ that have not yet been 
  visited are added to $N$, and $P~v$ is set to $u$. If $C$ becomes empty, $C$ and $N$ are swapped. If the target node is encountered, the algorithm immediately terminates.
  Note that this implementation is a generalized version of the usual queue implementation of BFS: While a queue enforces a complete order on the encountered nodes, our 
  implementation only enforces an ordering between nodes on the current level and nodes on the next level.

  For the actual algorithm, we wrap this algorithm by a procedure to handle the special case where source and target nodes are the same, and to extract a shortest path 
  from $P$ upon termination. Moreover, we implement the abstract specification of adding the successor nodes of a node by a loop.

  Finally, we prove the following theorem: %TODO: Not actually true, the impls are oly done in bfs2!
%     theorem
%       assumes "src∈V" 
%       shows "bfs src dst ≤ (spec p. 
%         case p of None ⇒ ¬conn src dst | Some p ⇒ shortestPath src p dst)"

@@

  Down to executable code:
    Implement things that where not yet specified/specified in a non-executable way:

      Residual graph: All operations of the algorithm can also be expressed as only modifying the residual graph. 
      Although we can compute the edges of the residual graph from the capacities and the current flow on the fly,
      this still requires 2 or 3 accesses to adjacency matrices per edge.

      Thus, we refine the algorithm to directly operate on residual graphs. The shortest path specification and bottleneck computation are already 
      phrased on residual graphs. What remains is the computation of the initial residual graph (which equals the Network),
      the augmentation of the residual graph, given an augmenting path and its bottleneck capacity, and the computation of the 
      resulting maximum flow from the residual graph, at the end of the algorithm.

      In order to refine these operations, we note that the residual graph uniquely determines the flow (and vice versa). 
%         definition (in Network) "flow_of_cf cf e ≡ (if (e∈E) then c e - cf e else 0)"

      For a valid flow $f$, we have that
%         lemma (in NFlow)
%           "flow_of_cf (residualGraph f) = f"
%           by (rule fo_rg_inv)

      Thus, augmentation can be expressed directly on the residual graph:
%         "residualGraph (augment (augmentingFlow p)) (u,v) = (
%           if (u,v)∈set p then (residualGraph f (u,v) - bottleNeck p)
%           else if (v,u)∈set p then (residualGraph f (u,v) + bottleNeck p)
%           else residualGraph f (u,v))"

      Having these characterizations, a straightforward refinement yields an algorithm that operates on residual graphs.

      In a next step, we have to specify iterative implementations for finding the bottleneck capacity and doing the augmentation.
      This is straightforward by folding over the path, exploiting that an augmenting path is simple, and thus never contains an edge 
      and its flipped edge at the same time.

      By combination of c and f, and pred-succ. TODO: Directly!
        All we need is successor function. Implemented by tabulating 
          adjacent nodes of each node (pred-succ), and using c and f to filter out actual successor nodes. 
      augpath-spec by BFS/DFS (which work on succ-functions)
      bottleneck and augment: Give iterative impl.

      In order to find an augmenting path, the BFS algorithm has to compute the successors of a node. 
      Although this can be implemented on the capacity matrix by iterating over all nodes, this implementation tends to be inefficient for sparser graphs.
      A common optimization is to pre-compute a map from nodes to adjacent nodes in the network. As an edge in the residual graph is either parallel or reverse to 
      a network edge, it is enough to iterate over the adjacent nodes in the network, and check whether they are actually successors in the residual graph.
      It is straightforward to show that this implementation actually returns the successors in the residual graph:
%         lemma (in RGraph) rg_succ_ref:
%           assumes A: "is_pred_succ ps c"
%           assumes B: "u∈V"
%           shows "rg_succ2 ps cf u ≤ SPEC (λl. (l,cf.E``{u}) ∈ ⟨Id⟩list_set_rel)"

    Use efficient data structures:
      Now we have an abstract version of the algorithm that only relies on executable concepts.
      However, to yield an efficient algorithm, we have to choose efficient data structures.

      We assume that the nodes are natural numbers less than $N$. 
      Then we implement the capacity matrix of the residual graph by an array with row-major indexing. 
      The adjacency map for network nodes is implemented as an array of lists. 
      In the BFS algorithm, we use two lists, one that holds the nodes of the current level, and another 
      in which nodes of the next level are collected. The predecessor map is implemented as an array.

      The input of the algorithm is still a function mapping network edges to capacities, which is tabulated once 
      to obtain the initial residual graph. This ensures some flexibility in using the algorithm, without preventing 
      efficient implementations (The function can be based on array lookup itself\footnote{Due to technical limitations of our tools, this function 
        cannot depend on heap content itself, but it can use the efficient functional and pseudo-functional datastructures provided by the Collection Framework.}).

      The output of the algorithm is just the residual graph. The user can decide how he computes the maximum flow from it. For example,
      in order to compute the maximum flow value, only the outgoing edges of the source node have to be computed, which is typically less 
      expensive than computing the complete flow matrix. The correctness theorem of the algorithm abstractly states how to obtain the maximum flow from the output.

      The last refinement step was performed using the Sepref tool, which supports data refinements from functional to imperative algorithms in Isabelle/HOL.
      [Few sentences on Sepref]

      Arrays for c,f, and ps. c and f in row-major indexing.

@@

\section{Benchmarking}
  What implementations did we compare: Authors? Where do implementations come from? 
  [We must convince the reader that we did not intentionally chose bad implementations to compare us against!]

  Comparsion of algorithms (Modified data structure: Flow vs Flow-like vs ResGraph)
    Computing successors in BFS: Filter by visited set first.
      -> Technically more challenging, when computing on flow: 1 access first vs 2 accesses first. On resGraph: No advantage in memory accesses (However, plain array access vs. matrix access)

\section{Conclusion} 
  ... and related work
    Mizar-Formalization: What did they do.
    Future Work: Dinic. (Actually, our abstract scheme [almost] covers Dinic's algorithm!)

@@

  Formal proof of mincut maxflow
  fofo-scheme
  inst to edmonds karp
  complexity analysis of edmonds karp
  refinement down to executable code. Roughly 2.5 times slower than Java. (What about OCaml)
    + NetCheck
